\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-11-09 k-planar.tex
Pfft

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DFMS21} use it to prove that planar graphs have $p$-centered chromatic number $O(p^2\log p)$ and give a matching lower bound.
\item \citet{DEJGMM21} use it to find asymptotically optimal adjacency labellings of planar graphs (resolving a problem of \citet{kannan.naor.ea:implicit}).
\item \citet{EJM} use it to show the existence of a `universal graph' with $n^{1+o(1)}$ vertices and edges that contains every $n$-vertex planar graph as an induced subgraph (resolving a problem of \citet{babai.chung.ea:on}).
\end{compactitem}
In addition, \cref{PlanarProduct} has been used to resolve or make substantial progress on a number of other problems on planar graphs, including $\ell$-vertex ranking~\citep{BDJM}, twin-width~\citep{BKW}, and odd colourings~\citep{DMO}.

@@

We start with the example of $k$-planar graphs. A graph is \defin{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings, where no three edges cross at a single point (see \cref{k_planar_section} for a formal definition). Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. It is well-known that the family of $k$-planar graphs is not minor-closed.  Indeed, 1-planar graphs may contain arbitrarily large complete graph minors~\citep{dujmovic.eppstein.ea:structure}. Hence the above results are not applicable for  $k$-planar graphs. We extend \cref{PlanarProduct} as follows.

@@

The above result for $k$-planar graphs (\cref{kPlanarProduct}) is in fact a special case of a more general result that relies on the following definition. A non-empty set $\SS$ of non-trivial paths in a graph $G$ is a \defin{$(k,d)$-shortcut system} (for $G$) if:

@@

\begin{table}[h]
  \centering{
    \begin{tabular}{lccl}
      \hline
      {Graph class} & $\tw(H)$ & $\ell$ & {Reference} \\ \hline
      planar & $3^*$ & $3$ & \cite{DJMMUW20} \\
      planar & $4^*$ & $2$ & \cite{bose.morin.ea:optimal} \\
      planar & $6^*$ & $1$ & \cite{UWY22} \\
      genus $g$ & $3^*$ & $\max\{2g,3\}$ & \cite{DHHW22} \\
      genus $g$ & $2g+6$ & 1 & \cite{UWY22} \\
      apex-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      % B-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      $k$-planar & $\binom{k+4}{3}-1$ & $18k^2 + 48k + 30$ & \cref{kPlanarProduct} \\
      $(g,k)$-planar & $\binom{k+4}{3}-1$ & $\max\{2g,3\}(6k^2+16k+10)$ & \cref{gkPlanarProduct} \\
      $(g,\delta)$-string & $\binom{\delta+4}{3}-1$ & $\max\{2g,3\}(\delta^4 + 4\delta^3 + 9\delta^2 + 10\delta+4)$ & \cref{StringProduct} \\
      $k$-nearest-neighbour & $O(k^6)$ & $O(k^4)$ & \cref{nn_product_structure} \\
      $d$-framed & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{d_framed_product_stucture} \\
      $1$-planar & $3^*$ & $7$ & \cref{1_planar_product} \\
      $d$-map & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{dMapProduct} \\
      $(g,d)$-map & $9$ & $\max\{2g,3\}\,(7d^2 -21d)$ & \cref{gdMapProduct} \\ \hline
      \multicolumn{4}{l}{$^*$ these bounds also apply to the simple treewidth of $H$.}
    \end{tabular}
  }
  \caption{Product structure theorems (of the form $G\subseteq H\boxtimes P\boxtimes K_\ell$).}
  \label{summary_table}
\end{table}

The remainder of this paper is organized as follows:
\begin{compactitem}
  \item In \cref{Structure} we prove our main result for shortcut systems (\cref{ShortcutProduct}), and its optimisations for $k$-planar graphs and their genus-$g$ generalisation (\cref{kPlanarProduct,gkPlanarProduct}).
  \item \Cref{examples} uses \cref{ShortcutProduct,kPlanarProduct,gkPlanarProduct} to derive product structure theorems for bounded-degree string graphs, powers of bounded-degree graphs, map graphs, and $k$-nearest neighbour graphs.
  \item \Cref{FramedSection} proves a product structure theorem for $d$-framed multigraphs in which the graph $H$ has treewidth $3$ and is planar.  Using this result, we derive improved product structure theorems for $1$-planar graphs and planar map graphs, again with a planar graph $H$.
  \item \Cref{Applications} quickly surveys applications of product structure theorems and the consequences of the current work.
\end{compactitem}

@@

We now prove \cref{ShortcutPartition}.
For convenience, it will be helpful to assume that $\SS$ contains a length-1 $vw$-shortcut for every edge $vw\in E(G)$.  Since $G^\SS$ is defined to be a supergraph of $G$, this assumption has no effect on $G^{\SS}$ but eliminates special cases in some of our proofs.  For a given $H$-partition $\mathcal{H}:=\{B_x:x\in V(H)\}$ of a graph $G$, we are frequently interested in the unique vertex $x\in V(H)$ such that $B_x$ contains a particular vertex $v\in V(G)$.  We will write this concisely by writing that ``$P_x$ is the part in $\mathcal{H}$ that contains $v$.''

Let $T$ be a tree rooted at some node $x_0\in V(T)$.  A node $a\in V(T)$ is a \defin{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \defin{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in V(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \defin{strict} $T$-ancestor of $x$ if $a\neq x$.
The \defin{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.  The \defin{lowest common $T$-ancestor} of a non-empty set $S\subseteq V(T)$ is the $T$-ancestor of every node in $S$ with maximum $T$-depth. In the following, we treat any subtree $T'$ of $T$ as a rooted tree whose root is the lowest common $T$-ancestor of $V(T')$.

@@

  Begin with any width-$t$ tree-decomposition $(B_x:x\in V(T_0))$ of $H$ that uses some tree $T_0$ and has no empty bags.  Select any node $x_1\in V(T_0)$, add a leaf $x_0$, with $B_{x_0}=\emptyset$, adjacent to $x_1$ and root $T_0$ at $x_0$. (The purpose of $x_0$ is to ensure that every node $x$ for which $B_x$ is non-empty has a parent.)  Let $f:V(H)\to V(T)$ be the function that maps each $x\in V(H)$ onto the root of the subtree $T_0[x]:=T_0[\{y\in V(T_0): x\in B_y\}]$.  If $f$ is not one-to-one, then select some distinct pair $x,y\in V(H)$ with $a:=f(x)=f(y)$.  Subdivide the edge between $a$ and its $T$-parent by introducing a new node $a'$ with $B_{a'}=B_{a}\setminus\{x\}$. Now $f(y)=a'$ and $f(x)=a$, so this modification reduces the number of distinct pairs $x,y\in V(H)$ with $f(x)=f(y)$.  Repeatedly performing this modification will eventually produce a tree-decomposition $(B_x:x\in V(T_0))$ of $H$ in which $f$ is one-to-one.

@@

  Since $H_v$ contains a path from $x_0$ to $x$ for each $x\in V(H_v)\setminus\{x_0\}$,  $H_v$ is connected.  Since $H_v$ is connected, $x_r$---which is the lowest common $T$-ancestor of $V(H_v)$---is itself a node of $V(H_v)$.  Therefore $H_v$ contains a path $x_0,\ldots,x_r$ of length most $k-1$.  By the definition of $H_v$, $v$ participates in each of $x_0,\ldots,x_r$.
\end{proof}

@@

\begin{proof}
  % Recall that $S_x$ is defined by vertices that participate in $x$, and these are vertices that are either in $Y_x$ or in shortcuts that cross $Y_x$.
  It follows from \cref{short_path_in_h} that, for each $v\in V(H)$, $v$ participates in $x:=a(v)$.  Therefore, the contents of $S_x$ are limited to elements that are in $Y_x$ or vertices in shortcuts that cross $x$.  We say that a vertex $w\in Y_x$ \defin{contributes} a vertex $v\in S_x$ if $v=w$ or if some path in $\SS$ that contains $v$ has $w$ as an internal vertex.  We upper bound the number of vertices in $S_x\cap L_i$ by upper-bounding the number of vertices contributed to $S_x\cap L_i$ by each $w\in Y_x$.

@@

\begin{proof}
  For the tree-decomposition $(C_x:x\in V(T))$ of $J$ we use the same tree $T$ used in the tree-decomposition $(B_x:x\in V(T))$ of $H$. For each node $x$ of $T$, we define $C_x$ as follows: $C_x$ contains every $T$-ancestor $a$ of $x$ such that $J$ contains an edge $ax'$ where $x$ is a $T$-ancestor of $x'$ (including the possibility that $x=x'$).  Additionally, $C_x$ includes $x$ if $x\in V(J)$.
  \cref{i-ancestor} ensures that, for every edge $ax'\in E(J)$, both $a$ and $x'$ are contained in $C_{x'}$.  The connectivity of $T[a]:=T[\{x\in V(T):a\in C_x\}]$ follows from the fact that, for every node $x'\in V(T[a])$, every node $x$ on the path in $T$ from $a$ to  $x'$ is also a node of $T[a]$. Therefore $(C_x:x\in V(T))$ is indeed a tree-decomposition of $J$.

@@

  Let $H^+$ denote the supergraph of $H$ that contains the edge $xy$ if and only if some bag $B_z\in\mathcal{T}$ contains both $x$ and $y$ (that is, $H^+$ is the chordal closure of $H$ with respect to $\mathcal{T}$).  Let $\overrightarrow{H}^+$ be the directed graph obtained by directing each edge $xy$ of $H^+$ in the direction $\overrightarrow{xy}$ so that $y$ is a $T$-ancestor of $x$. Observe that any directed path in $\overrightarrow{H}^+$ that begins at $x$ leads to a $T$-ancestor of $x$.  \citet[Lemma~13]{PS21} show that, for any $x\in V(\overrightarrow{H}^+)$, the number of strict $T$-ancestors of $x$ that can be reached by a directed path of length at most $k$ that begins at $x$ is at most $\binom{k+t}{t}-1$.

@@

    \item $x_r$ is a strict $T$-descendant of $x$.
      Since  $y_0=x_r$ is a strict $T$-descendant of $x$ and $y_s$ is a $T$-ancestor of $x$, the path in $T$ from $y_s$ to $y_0$ includes $x$.  Since $H[\{y_0,\ldots,y_s\}]$ is connected, at least one of $y_0,\ldots,y_s$, say $y_i$, is contained in $B_x$. Therefore $xy_i$ is an edge of $H^{+}$.  Therefore $x,y_i,\ldots,y_s$ is a path in $H^+$ from $x$ to $a$ of length at most $s+1\le k$ and $v_r$ participates in each of $y_i,\ldots,y_s$.
  \end{enumerate}
  In each case, we obtain a path $z_0,\ldots,z_{s'}$ in $H$ of length at most $k$ from $z_0:=x$ to $z_{s'}:=a$ such that $v_r$ participates in each of $z_1,\ldots,z_{s'}$.  By definition, $a$ is a $T$-ancestor of $x$.  Since $v_r$ participates in each of $z_1,\ldots,z_{s'}$, $a=a(v_r)$ is also a $T$-ancestor of each of $z_1,\ldots,z_{s'}$.  Let $W$ be the sequence obtained from $z_0,\ldots,z_{s'}$ by replacing $z_i$ with the lowest common $T$-ancestor of $z_0,\ldots,z_i$, for each $i\in\{0,\ldots,s'\}$.  Then $W$ is a lazy walk in $H^+$ of length at most $k$ that begins at $z_0=x$ and ends at $z_{s'}=a$. By removing duplicates from $W$ we obtain a path in $H^+$ that is also a directed path in $\overrightarrow{H}^+$.  This completes the proof of \cref{general-bag-size}.
\end{proof}

@@

%%%%%%%
\subsection{String Graphs}
\label{first_example}

@@

\subsection{Map Graphs}

@@

The following definition is a convenient way to establish the planarity of $H$ in \cref{d_framed_product_stucture}:  A $T$-decomposition of a graph $G$ is \defin{$t$-simple} if each of its bags has size at most $t+1$ and, for each $t$-element subset $\{v_1,\ldots,v_t\}\subseteq V(G)$, at most two bags contain $v_1,\ldots,v_t$.  To show the planarity of $H$ in \cref{d_framed_product_stucture} we use the fact that a graph has a 3-simple tree-decomposition if and only if it has treewidth at most 3 and is planar~\citep{kratochvil.vaner:note}.

>>>>

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DFMS21} use it to prove that planar graphs have $p$-centered chromatic number $O(p^2\log p)$ and give a matching lower bound.
\item \citet{DEJGMM21} use it to find asymptotically optimal adjacency labellings of planar graphs (resolving a problem of \citet{kannan.naor.ea:implicit}).
\item \citet{EJM} use it to show the existence of a `universal graph' with $n^{1+o(1)}$ vertices and edges that contains every $n$-vertex planar graph as an induced subgraph (resolving a problem of \citet{babai.chung.ea:on}).
\end{compactitem}
In addition, \cref{PlanarProduct} has been used to resolve or make substantial progress on a number of other problems on planar graphs, including $\ell$-vertex ranking~\citep{BDJM} and twin-width~\citep{BKW}.

@@

We start with the example of $k$-planar graphs. A graph is \defin{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings, where no three edges cross at a single point (see \cref{k_planar_section} for a formal definition). Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. It is well-known that the family of $k$-planar graphs is not minor-closed.  Indeed, 1-planar graphs may contain arbitrarily large complete graph minors~\citep{dujmovic.eppstein.ea:structure}. Hence the preceding results are not applicable for  $k$-planar graphs. We extend \cref{PlanarProduct} as follows.

@@

The preceding result for $k$-planar graphs (\cref{kPlanarProduct}) is in fact a special case of a more general result that relies on the following definition. A non-empty set $\SS$ of non-trivial paths in a graph $G$ is a \defin{$(k,d)$-shortcut system} (for $G$) if:

@@

\begin{table}[h]
  \centering{
    \begin{tabular}{lccl}
      \hline
      {Graph class} & $\tw(H)$ & $\ell$ & {Reference} \\ \hline
      planar & $3^*$ & $3$ & \cite{DJMMUW20} \\
      planar & $4^*$ & $2$ & \cite{bose.morin.ea:optimal} \\
      planar & $6^*$ & $1$ & \cite{UWY22} \\
      genus $g$ & $3^*$ & $\max\{2g,3\}$ & \cite{DHHW22} \\
      genus $g$ & $2g+6$ & 1 & \cite{UWY22} \\
      apex-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      % B-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      $k$-planar & $\binom{k+4}{3}-1$ & $18k^2 + 48k + 30$ & \cref{kPlanarProduct} \\
      $(g,k)$-planar & $\binom{k+4}{3}-1$ & $\max\{2g,3\}(6k^2+16k+10)$ & \cref{gkPlanarProduct} \\
      $(g,\delta)$-string & $\binom{\delta+4}{3}-1$ & $\max\{2g,3\}(\delta^4 + 4\delta^3 + 9\delta^2 + 10\delta+4)$ & \cref{StringProduct} \\
      $k$-nearest-neighbour & $O(k^6)$ & $O(k^4)$ & \cref{nn_product_structure} \\
      $d$-framed & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{d_framed_product_stucture} \\
      $1$-planar & $3^*$ & $7$ & \cref{1_planar_product} \\
      $d$-map & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{dMapProduct} \\
      $(g,d)$-map & $9$ & $\max\{2g,3\}\,(7d^2 -21d)$ & \cref{gdMapProduct} \\ \hline
      \multicolumn{4}{l}{$^*$ these bounds also apply to the simple treewidth \cite{knauer.ueckerdt:simple} of $H$.}
    \end{tabular}
  }
  \caption{Product structure theorems (of the form $G\subseteq H\boxtimes P\boxtimes K_\ell$).}
  \label{summary_table}
\end{table}

The remainder of this paper is organized as follows:
\begin{compactitem}
  \item In \cref{Structure} we prove our main result for shortcut systems (\cref{ShortcutProduct}), and its optimisations for $k$-planar graphs and their genus-$g$ generalisation (\cref{kPlanarProduct,gkPlanarProduct}).
  \item \Cref{examples} uses \cref{ShortcutProduct,kPlanarProduct,gkPlanarProduct} to derive product structure theorems for bounded-degree string graphs, powers of bounded-degree graphs, map graphs of bounded clique number, and $k$-nearest neighbour graphs.
  \item \Cref{FramedSection} proves a product structure theorem for $d$-framed multigraphs in which the graph $H$ has treewidth $3$ and is planar.  Using this result, we derive improved product structure theorems for $1$-planar graphs and planar map graphs, again with a planar graph $H$.
  \item \Cref{Applications} quickly surveys applications of product structure theorems and the consequences of the current work.
\end{compactitem}

@@

We now prove \cref{ShortcutPartition}.
For convenience, it will be helpful to assume that $\SS$ contains a length-1 $vw$-shortcut for every edge $vw\in E(G)$.  Since $G^\SS$ is defined to be a supergraph of $G$, this assumption has no effect on $G^{\SS}$ but eliminates special cases in some of our proofs.  For a given $H$-partition $\mathcal{H}:=\{B_x:x\in V(H)\}$ of a graph $G$, we are frequently interested in the unique vertex $x\in V(H)$ such that $B_x$ contains a particular vertex $v\in V(G)$.  We express this concisely by writing that ``$P_x$ is the part in $\mathcal{H}$ that contains $v$.''

Let $T$ be a tree rooted at some node $x_0\in V(T)$.  A node $a\in V(T)$ is a \defin{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \defin{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in V(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \defin{strict} $T$-ancestor of $x$ if $a\neq x$.
The \defin{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.  The \defin{lowest common $T$-ancestor} of a non-empty set $S\subseteq V(T)$ is the maximum $T$-depth node $a\in V(T)$ that is a $T$-ancestor of every node in $S$.  In the following, we treat any subtree $T'$ of $T$ as a rooted tree whose root is the lowest common $T$-ancestor of $V(T')$.

@@

  Begin with any width-$t$ tree-decomposition $(B_x:x\in V(T_0))$ of $H$ that uses some tree $T_0$ and has no empty bags.  Select any node $x_1\in V(T_0)$, add a leaf $x_0$, with $B_{x_0}=\emptyset$, adjacent to $x_1$ and root $T_0$ at $x_0$. (The sole purpose of $x_0$ is to ensure that every node $x$ for which $B_x$ is non-empty has a parent.)  Let $f:V(H)\to V(T)$ be the function that maps each $x\in V(H)$ onto the root of the subtree $T_0[x]:=T_0[\{y\in V(T_0): x\in B_y\}]$.  If $f$ is not one-to-one, then select some distinct pair $x,y\in V(H)$ with $a:=f(x)=f(y)$.  Subdivide the edge between $a$ and its $T$-parent by introducing a new node $a'$ with $B_{a'}=B_{a}\setminus\{x\}$. Now $f(y)=a'$ and $f(x)=a$, so this modification reduces the number of distinct pairs $x,y\in V(H)$ with $f(x)=f(y)$.  Repeatedly performing this modification will eventually produce a tree-decomposition $(B_x:x\in V(T_0))$ of $H$ in which $f$ is one-to-one.

@@

  Since $H_v$ contains a path from $x_0$ to $x$ for each $x\in V(H_v)\setminus\{x_0\}$,  $H_v$ is connected.  Since $H_v$ is connected, $x_r$---which is the lowest common $T$-ancestor of $V(H_v)$---is itself a node of $V(H_v)$.  Therefore $H_v$ contains a path $x_0,\ldots,x_r$ of length at most $k-1$.  By the definition of $H_v$, $v$ participates in each of $x_0,\ldots,x_r$.
\end{proof}

@@

\begin{proof}
  % Recall that $S_x$ is defined by vertices that participate in $x$, and these are vertices that are either in $Y_x$ or in shortcuts that cross $Y_x$.
  It follows from \cref{short_path_in_h} that, for each $v\in V(H)$, $v$ participates in $x:=a(v)$.  Therefore, for each $x\in V(J)$, the contents of $S_x$ are limited to vertices that are in $Y_x$ or in shortcuts that cross $x$.  We say that a vertex $w\in Y_x$ \defin{contributes} a vertex $v\in S_x$ if $v=w$ or if some path in $\SS$ that contains $v$ has $w$ as an internal vertex.  We upper bound the number of vertices in $S_x\cap L_i$ by upper-bounding the number of vertices contributed to $S_x\cap L_i$ by each $w\in Y_x$.

@@

\begin{proof}
  For the tree-decomposition $(C_x:x\in V(T))$ of $J$ we use the same tree $T$ used in the tree-decomposition $(B_x:x\in V(T))$ of $H$. For each node $x$ of $T$, we define $C_x$ as follows: $C_x$ contains every $T$-ancestor $a$ of $x$ such that $J$ contains an edge $ax'$ where $x$ is a $T$-ancestor of $x'$ (including the possibility that $x=x'$).  Additionally, $C_x$ includes $x$ if $x\in V(J)$.
  \cref{i-ancestor} ensures that, for each edge $ax'\in E(J)$, both $a$ and $x'$ are contained in $C_{x'}$.  The connectivity of $T[a]:=T[\{x\in V(T):a\in C_x\}]$ follows from the fact that, for every node $x'\in V(T[a])$, every node $x$ on the path in $T$ from $a$ to  $x'$ is also a node of $T[a]$. Therefore $(C_x:x\in V(T))$ is indeed a tree-decomposition of $J$.

@@

  Let $H^+$ denote the supergraph of $H$ that contains the edge $xy$ if and only if some bag $B_z\in\mathcal{T}$ contains both $x$ and $y$ (that is, $H^+$ is the chordal closure of $H$ with respect to $\mathcal{T}$).  Let $\overrightarrow{H}^+$ be the directed graph obtained by directing each edge $xy$ of $H^+$ in the direction $\overrightarrow{xy}$ so that $y$ is a $T$-ancestor of $x$. Observe that any directed path in $\overrightarrow{H}^+$ that begins at $x$ leads to a $T$-ancestor of $x$.  \citet[Lemma~13]{PS21} show that, for any $x\in V(\overrightarrow{H}^+)$, the number of strict $T$-ancestors of $x$ that can be reached by a directed path in $\overrightarrow{H}^+$ of length at most $k$ that begins at $x$ is at most $\binom{k+t}{t}-1$.

@@

    \item $x_r$ is a strict $T$-descendant of $x$.
      Since  $y_0=x_r$ is a strict $T$-descendant of $x$ and $y_s=a$ is a $T$-ancestor of $x$, the path in $T$ from $y_0$ to $y_s$ includes $x$.  Since $H[\{y_0,\ldots,y_s\}]$ is connected, at least one of $y_0,\ldots,y_s$, say $y_i$, is contained in $B_x$. Therefore $xy_i$ is an edge of $H^{+}$.  Therefore $x,y_i,\ldots,y_s$ is a path in $H^+$ from $x$ to $a$ of length at most $s+1\le k$ and $v_r$ participates in each of $y_i,\ldots,y_s$.
  \end{enumerate}
  In each case, we obtain a path $z_0,\ldots,z_{s'}$ in $H$ of length $s'\le k$ from $z_0:=x$ to $z_{s'}:=a$ such that $v_r$ participates in each of $z_1,\ldots,z_{s'}$.  By definition, $a$ is a $T$-ancestor of $x$.  Since $v_r$ participates in each of $z_1,\ldots,z_{s'}$, $a=a(v_r)$ is also a $T$-ancestor of each of $z_1,\ldots,z_{s'}$.  Let $W$ be the sequence obtained from $z_0,\ldots,z_{s'}$ by replacing $z_i$ with the lowest common $T$-ancestor of $z_0,\ldots,z_i$, for each $i\in\{0,\ldots,s'\}$.  Then $W$ is a lazy walk in $H^+$ of length at most $k$ that begins at $z_0=x$ and ends at $z_{s'}=a$. By removing duplicates from $W$ we obtain a path in $H^+$ that is also a directed path in $\overrightarrow{H}^+$.  This completes the proof of \cref{general-bag-size}.
\end{proof}

@@

%%%%%%%
\subsection{Bounded-Degree String Graphs}
\label{first_example}

@@

\subsection{Map Graphs of Bounded Clique Number}

\note{PM}{To make the referee happy I've added the ``bounded-degree'' qualifier to string graphs and the ``bounded clique number'' qualifier to map graphs.  Do we want to explcitly state that the clique number of a $(g,d)$-map graph is bounded by a function of $g$ and $d$?  Proof (for the $g=0$ case): A $d$-map representation gives a drawing with at most $\floor{\tfrac{d-2}{2}}\ceil{\tfrac{d-2}{2}}\in O(d^2)$ crossings per edge, but any drawing of $K_{\omega}$ on a surface of genus-$g$ has an edge that is crossed $\Omega(\omega^2)$ times. So if $G$ is a $d$-map graph then $\omega(G) \in O(d)$.}

@@

The following definition is a convenient way to establish the planarity of $H$ in \cref{d_framed_product_stucture}:  A $T$-decomposition of a graph $G$ is \defin{$t$-simple} if each of its bags has size at most $t+1$ and, for each $t$-element subset $\{v_1,\ldots,v_t\}\subseteq V(G)$, at most two bags contain $v_1,\ldots,v_t$.  To show the planarity of $H$ in \cref{d_framed_product_stucture} we use the fact that a graph has a 3-simple tree-decomposition if and only if it has treewidth at most 3 and is planar~\citep{knauer.ueckerdt:simple,kratochvil.vaner:note}.
